Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-16 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Built and incorporated new FDS images
Completed spell check through Ch 8

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 mega-pixel resolution.  A mini DV camcorder with 0.7 mega-pixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

For the purposes of this study, a modified version of the test apparatus was used.  The radiant panel and the controls for the radiant panel remained in place, as documented in ASTM~E162. However, the sample mounting system was changed. A rail and trolley system was developed by Lawson and Twilley to expose fire fighter protective equipment to various heat fluxes by being able to move the test specimen to a predetermined position that correlated with a given heat flux~\cite{Lawson:1999}.  This trolley was used to hold a sample of painted gypsum wallboard that was 362~mm square.  The samples were vertical and parallel to the radiant panel. Figure~\ref{Rad_Panel} has photographs which show the radiant panel test arrangement.

@@

Table~\ref{tab:Point_Source_HFPUF} displays a similar development based on the average peak heat release rate, 45~kW, for the polyurethane foam burner.
Similar to the results from the gasoline comparison, the predictive heat flux values at all of the burner positions are low relative to the experimental data.  But again the results will be examined along a more qualitative basis. The experimental data provides an average peak value of 25~kW/m$^2$ for $R=0.31~m$. For the polyurethane foam burns the peak values were only averaged over a 15 second range due to the brief nature of the peak heat release rate. At $R=0.31~m$ there was no charring of the painted paper on the gypsum wallboard, only soot deposition.  No thermal damage.  At $R=0.23~m$ or less, the heat flux was high enough to indicate ignition in the experiments and that was the result.  Only the prediction at $R=0.15~m$ would definitely indicate that the gypsum wallboard would ignite based on the threhold heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	     & Heat Flux (kW/m$^2$)     &  Polyurethane (kW/m$^2$)    	\\ \hline \hline
		45       &   .35 		& 0.76 	     & 2.1 $\pm$ 0.17 (8~\%) 	&  3.5 $\pm$ 0.28 (8~\%) 	   	 \\
		45       &   .35    	& 0.46	     & 6.0 $\pm$ 0.48 (8~\%) 	&  12.3 $\pm$ 0.98 (8~\%)  	     \\
		45       &   .35 		& 0.31	     & 13.5 $\pm$ 1.1 (8~\%) 	&  25 $\pm$ 2.0 (8~\%)   	      \\
        45       &   .35 		& 0.23	     & 24.1 $\pm$ 1.9 (8~\%) 	&  36 $\pm$ 2.9 (8~\%)		      \\
        45       &   .35 		& 0.15	     & 53.5 $\pm$ 4.3 (8~\%) 	&  59 $\pm$ 4.7 (8~\%)   	       \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 45~kW) with heat flux data from the instrumented gypsum boardwall experiments]{Comparison of calculated point source values with heat flux data from the polyurethane foam burner instrumented gypsum boardwall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Pitts:2006}.}
	\label{tab:Point_Source_HFPUF}
\end{table}

This was an interesting exercise but it was not very definitive.  If we were to attempt to improve the predictions by using radiative fractions that are known to be representative of the fuels, the predictions in some cases might get worse.  For example, recommended values in the literature for the $chi_r$ of natural gas range from 0.15 to 0.25 for similar sized burners~\cite{Hostikka:3,McCaffrey:1982}.  The default radiative fraction of 0.35 is inline with the values found in the literature for gasoline~\cite{Hamins:1991,McGrattan:2000}.  Literature values for polyurethane foam were elusive.

@@

\begin{table}
\centering
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
Method          &   \multicolumn{3}{|c|}{Flame Heights (m)} \\ \cline{2-4}
                &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Photographic    &       0.70 $\pm$ 0.23 (33~\%)   &	0.84 $\pm$ 0.28 (33~\%) 	& 0.46 $\pm$ 0.23 (50~\%)   \\
Video Based     &   	0.70 $\pm$ 0.11 (15~\%)   &	0.70 $\pm$ 0.1 (14~\%) 	& 0.47 $\pm$ 0.16 (34~\%)   \\
Fire Pattern    &   	0.74 $\pm$ 0.12 (16~\%)   &	0.83 $\pm$ 0.15 (18~\%) 	& 0.24 $\pm$ 0.12 (50~\%)   \\
Heskestad (range)  &   	0.94 to 1.06		    &	0.65 to 0.93        	& 0.36 to 0.72   \\
\hline
\end{tabular}
 \caption[Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions]{Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions. The measured results are shown with 95~\% confidence limits.  The Heskestad median flame height calculations are presented as a range based on the uncertainies of the average peak heat release rate of each fuel type.} 
 \label{tab:Comparing_Flame_Heights}
\end{table}

 The comparison for the natural gas burner shows agreement between the measured median flame height and the average maximum fire pattern height. However, the range of calculated median flame height values is significantly different from the mesured values. Based on the computational flame height assessment, Figure~\ref{Gas_and_Gasoline_Contours}, the Heskestad correlation provided a flame height value in the intermittent region where the flame is present less than 10~\% of the time.

The comparsion of the median flame height values, the average maximum fire pattern height, and the calculated flame height range are all in agreement for the gasoline fueled pan fires. This is significant because the 95~\% confidence limits for the fire pattern and the video based flame height are both less than 20~\%.

@@

Overall the Heskestad equation provided reasonable estimates of the median flame height.  The key to the successful use of any model regardles of the simplicity or complexity is using appropriate data.

@@

In this section, the natural gas burner will be modelled with a range of grid cell resolutions and radiative fraction values.  The results are compared against the source fire characterization data to determine which grid cell resolution and radiative fraction provided the best fire simulation for use in the simulations of the wall and compartment experiments.

@@

Choosing the size of the grid when performing an analysis with FDS is critical in order to get the level of resolution needed to adequately simulate the required details within a simulation. In studies where large rooms or portions of buildings have been modeled grid cells on the order of 100~mm have been used and resulted in the successful simultation of fire flow and fire growth trends ~\cite{Grosshandler:Station,Madrzykowski:2004,Vettori:2000, Barowy:2012,Overholt:San_Francisco,Weinschenk:Chicago}.

@@

To examime the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were choosen to model the fire. The resulting resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) resolution.  FDS simulations of the source fire characterization experiments will be conducted with these resolutions and compared with the results from the experiments.

@@

The computational domain of the model had the geometry of a cube with 2.4~m on a side.  The resulting number of computational cells for each case ranged from 27,000 for the 80~mm cells to 1,728,000 for the 20~mm cells. For the 80 and 40~mm computational cell scenarios, the cases were run as a single mesh.  For the 20~mm case, the computational domain was divided into 8 uniform meshes.  Each mesh was 2.4~m by 1.2~m and 0.6~m tall.  The time domain for each model was chosen to be 300 seconds, which is similar to the burn time fior each of the fuels.

@@

FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the source fire experiments.  The thermocouples were positioned in a vertical array, centered on the burner, beginning 0.1~m above the burner surface and spaced every 0.1~m upto 1.2~m above the burner surface.  A total heat flux gauge was also modelled to provide data at the 0.95 m from the centerline of the burner and 0.5 m above the top of the burner.  
In addition, FDS features devices that can not be physically implemented.  One such device, is used to determine the heat release rate per unit length.  This device was used to determine flame height~\cite{McDermott:2011}.

@@

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combustion zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Adjusting the FDS corresponding predictions from the RI=10 and RI=20 models down to the 0.3~m position appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explaination may be only one possibility.

These simulation results are consistant with the comparisons made to the McCaffrey's plume correlation from his study, `Purely Buoyant Diffusion Flames'~\cite{McCaffrey:1979}, in the FDS Validation Guide~\cite{FDS_Validation_Guide}. McCaffrey used a natural gas burner, 0.30~m square, which is similar in size to the burner used in this study.  McCaffrey noted that, "visually non-axisymmetric behavior could be observed only very close to the burner surface".  In the McCaffrey study, temperatures were measured with Type K thermocouples covered with a 1.0~mm stainless steel sheath. The peak heat release rate in McCaffery's plume study was 57.5~kW, this heat release rate was also the closest to the 80~kW fires in the current study.  The comparison between the McCaffrey plume correlation and the FDS simulations at RI values of 5, 10 and 20 are shown in Figure~\ref{McCaffrey_Plume_comp}. Focusing on the data between the burner surface and the 1~m above the burner surface, FDS underpredicted the temperature near the surface of the burner and then overpredicted as the elevation increased to approximately 0.4~m and 1.3~m above the burner for the RI=10 and 20 cases respectively. Once the temperature comparisons were above the continuous and intermittant flame regions, the agreement between the cases and the correlation began to improve.  Similar to the current study, increased RIs did not improve the simulation of the vertical centerline temperatures close to the burner.

@@

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved argeement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead may be generating the differences between the gas temperature and the measured temperature FDS values.

@@

Another examinination of the plume temperature predictions was made to examine the impact of the range of radiative fractions on the FDS results.  The comparisons were made using the course grid, RI=5, and radiative fraction values of 0.20, 0.22, 0.25, and 0.30, as before. The simulations with radiative fractions between 0.2 and 0.25 were within the 2~$\sigma$ limits of the temperature measurements at each elevation.

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenerio showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI= 5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

The next comparsions were made with the natural gas source fire at 4 different distances from an instrumented gypsum board wall.  Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).  The simulations will replicate the same distances.  The simulation included a vertical gypsum board wall, 1.2~m wide, that extends from the floor of the model up to the top of the domain.  The material properties of the wall are based solely on the properties of gypsum that were cited in Chapter 4.  The values that are compared in the section include: vertical centerline temperatures above the burner, wall temperatures, and heat flux values at the position of the wall surface.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.28 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500~$^\circ$C~\cite{Gross:1985}.  While FDS can accomodate changes in thermoconductivy and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

@@

 The temperatures measured by thermocouples positioned along the vertical centerline of the natural gas burner will be compared with FDS generated temperures from two simulations, one with RI-5 and one with RI=10.  The measurement points were at intervals of 0.2~m from 0.2~m to 2.2~m above the burner.  Figure~\ref{FDS_Plume_IWGB_comp} contains four graphs.  Each graph shows the same types of comparison, the only difference is the distance away from the gypsum wallboard.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.

@@

Figure~\ref{FDS_Wall_Temp_IWGB_comp} is analogous to Figure~\ref{FDS_Plume_IWGB_comp} except it compares the gypsum wallboard surface temperatures against the FDS simulated temperatures.  So while the FDS values for the plume over predicted the temperatures, the FDS values for the wall surface temperatures are underpredicting.  This is consistent with the flame and plume moving away from the plume centerline and toward the wall.  The agreement between the FDS results and the measurements improves as the burner moves closer to the wall, however once the burner is against the wall, the degree of agreement between the FDS results and the measured temperatures decreased.  At this position the FDS results are overpredicting the temperature. This differences at the wall are most distinct in the region where the flame impinged on the wall.  The agreement improves in the plume area above the flames.  In the comparisons with the source fire experiments, the FDS generated temperatures were underpredicting closer to the burner surface and the overpredicting in the flame region as a result of the `combustion' occuring at higher elevations above the burner.  If a similar phenomena is occuring here, the wall maybe exaggerating the effect.

@@

The instrumented gypsum wallboard had two vertical arrays of total heat flux gauges positioned at 0.2~m intervals.  Each array had six gauges installed from 0.2~m up through 1.2~m above the burner.  One array was installed in line with the center of the burner and the other array was installed on the right edge of the burner.  The comparisons between the measured heat flux and the FDS simulations are provided in Figure~\ref{FDS_HFCL_IWNG_comp} and Figure~\ref{FDS_HFEdge_IWNG_comp}.  For the centerline gauges the trends between the heat flux measurements and the FDS values are similar to the trends with the temperatures and are consistant with the flame moving or leaning toward the wall.  This resulted in higher heat flux measures relative to the FDS values at the three positions away from the wall. When the burner was positioned against the wall the FDS values and the measurements converged to a higher level of agreement.

@@

The FDS values significantly underpredict the measured heat flux values along the right edge of the burner.  The best agreement occurs at the burner positions away from the wall.  At this position, the fire pattern on the walls show that the lines of demarcation are occuring very close to the edge of the burner.  Again flame movement seems to to be the main cause of the difference that is accentuated by a measurement location that is very sensitive to heat flux and position.

@@

Reviewing the videos from a natural gas burner instrumented gypsum wallboard experiment with the burner positioned at 0.310~m from the wall, we can see examples of the flame movement.  In the top row of Figure~\ref{FDS_IWNG_Flame_comp} the three images exhibit points in time where the flame was moving toward the instrumented wall and away from the verticle array of thermocouples that are positioned above the center line of the burner.  The bottom row of images are from the same fire experiment looking at it from the front.  These images also show how the pulsations and movements of the flame change positions relative to the fixed positions of the thermocouples and heat flux gauges. 
Figure~\ref{FDS_IWNG_SMV_comp} has a frontview and a side view Smokeview image from the FDS RI=10 simulation with the burner positioned at 0.31~m from the wall.  This simulation is comparable to the experiment used for Figure~\ref{FDS_IWNG_Flame_comp}.  While the simultion does not demonstrate the amount of eddy generation or side to side movement from the front, the image from the side does show that FDS is simulating that the flame is moving toward the wall, although not to the extent that the actual flame did.

>>>>

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 mega-pixel resolution.  A mini DV camcorder with 0.7 mega-pixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 mega-pixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

For the purposes of this study, a modified version of the test apparatus was used.  The radiant panel and the controls for the radiant panel remained in place, as documented in ASTM~E 162. However, the sample mounting system was changed. A rail and trolley system was developed by Lawson and Twilley to expose fire fighter protective equipment to various heat fluxes by being able to move the test specimen to a predetermined position that correlated with a given heat flux~\cite{Lawson:1999}.  This trolley was used to hold a sample of painted gypsum wallboard that was 362~mm square.  The samples were vertical and parallel to the radiant panel. Figure~\ref{Rad_Panel} has photographs which show the radiant panel test arrangement.

@@

Table~\ref{tab:Point_Source_HFPUF} displays a similar development based on the average peak heat release rate, 45~kW, for the polyurethane foam burner.
Similar to the results from the gasoline comparison, the predictive heat flux values at all of the burner positions are low relative to the experimental data.  But again the results will be examined along a more qualitative basis. The experimental data provides an average peak value of 25~kW/m$^2$ for $R=0.31~m$. For the polyurethane foam burns the peak values were only averaged over a 15 second range due to the brief nature of the peak heat release rate. At $R=0.31~m$ there was no charring of the painted paper on the gypsum wallboard, only soot deposition.  No thermal damage.  At $R=0.23~m$ or less, the heat flux was high enough to indicate ignition in the experiments and that was the result.  Only the prediction at $R=0.15~m$ would definitely indicate that the gypsum wallboard would ignite based on the threshold heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	     & Heat Flux (kW/m$^2$)     &  Polyurethane (kW/m$^2$)    	\\ \hline \hline
		45       &   .35 		& 0.76 	     & 2.1 $\pm$ 0.17 (8~\%) 	&  3.5 $\pm$ 0.28 (8~\%) 	   	 \\
		45       &   .35    	& 0.46	     & 6.0 $\pm$ 0.48 (8~\%) 	&  12.3 $\pm$ 0.98 (8~\%)  	     \\
		45       &   .35 		& 0.31	     & 13.5 $\pm$ 1.1 (8~\%) 	&  25 $\pm$ 2.0 (8~\%)   	      \\
        45       &   .35 		& 0.23	     & 24.1 $\pm$ 1.9 (8~\%) 	&  36 $\pm$ 2.9 (8~\%)		      \\
        45       &   .35 		& 0.15	     & 53.5 $\pm$ 4.3 (8~\%) 	&  59 $\pm$ 4.7 (8~\%)   	       \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 45~kW) with heat flux data from the instrumented gypsum board wall experiments]{Comparison of calculated point source values with heat flux data from the polyurethane foam burner instrumented gypsum board wall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Pitts:2006}.}
	\label{tab:Point_Source_HFPUF}
\end{table}

This was an interesting exercise but it was not very definitive.  If we were to attempt to improve the predictions by using radiative fractions that are known to be representative of the fuels, the predictions in some cases might get worse.  For example, recommended values in the literature for the $chi_r$ of natural gas range from 0.15 to 0.25 for similar sized burners~\cite{Hostikka:3,McCaffrey:1982}.  The default radiative fraction of 0.35 is in line with the values found in the literature for gasoline~\cite{Hamins:1991,McGrattan:2000}.  Literature values for polyurethane foam were elusive.

@@

\begin{table}
\centering
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
Method          &   \multicolumn{3}{|c|}{Flame Heights (m)} \\ \cline{2-4}
                &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Photographic    &       0.70 $\pm$ 0.23 (33~\%)   &	0.84 $\pm$ 0.28 (33~\%) 	& 0.46 $\pm$ 0.23 (50~\%)   \\
Video Based     &   	0.70 $\pm$ 0.11 (15~\%)   &	0.70 $\pm$ 0.1 (14~\%) 	& 0.47 $\pm$ 0.16 (34~\%)   \\
Fire Pattern    &   	0.74 $\pm$ 0.12 (16~\%)   &	0.83 $\pm$ 0.15 (18~\%) 	& 0.24 $\pm$ 0.12 (50~\%)   \\
Heskestad (range)  &   	0.94 to 1.06		    &	0.65 to 0.93        	& 0.36 to 0.72   \\
\hline
\end{tabular}
 \caption[Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions]{Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions. The measured results are shown with 95~\% confidence limits.  The Heskestad median flame height calculations are presented as a range based on the uncertainties of the average peak heat release rate of each fuel type.} 
 \label{tab:Comparing_Flame_Heights}
\end{table}

 The comparison for the natural gas burner shows agreement between the measured median flame height and the average maximum fire pattern height. However, the range of calculated median flame height values is significantly different from the measured values. Based on the computational flame height assessment, Figure~\ref{Gas_and_Gasoline_Contours}, the Heskestad correlation provided a flame height value in the intermittent region where the flame is present less than 10~\% of the time.

The comparison of the median flame height values, the average maximum fire pattern height, and the calculated flame height range are all in agreement for the gasoline fueled pan fires. This is significant because the 95~\% confidence limits for the fire pattern and the video based flame height are both less than 20~\%.

@@

Overall the Heskestad equation provided reasonable estimates of the median flame height.  The key to the successful use of any model regardless of the simplicity or complexity is using appropriate data.

@@

In this section, the natural gas burner will be modeled with a range of grid cell resolutions and radiative fraction values.  The results are compared against the source fire characterization data to determine which grid cell resolution and radiative fraction provided the best fire simulation for use in the simulations of the wall and compartment experiments.

@@

Choosing the size of the grid when performing an analysis with FDS is critical in order to get the level of resolution needed to adequately simulate the required details within a simulation. In studies where large rooms or portions of buildings have been modeled grid cells on the order of 100~mm have been used and resulted in the successful simulation of fire flow and fire growth trends ~\cite{Grosshandler:Station,Madrzykowski:2004,Vettori:2000, Barowy:2012,Overholt:San_Francisco,Weinschenk:Chicago}.

@@

To examine the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were chosen to model the fire. The resulting resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) resolution.  FDS simulations of the source fire characterization experiments will be conducted with these resolutions and compared with the results from the experiments.

@@

The computational domain of the model had the geometry of a cube with 2.4~m on a side.  The resulting number of computational cells for each case ranged from 27,000 for the 80~mm cells to 1,728,000 for the 20~mm cells. For the 80 and 40~mm computational cell scenarios, the cases were run as a single mesh.  For the 20~mm case, the computational domain was divided into 8 uniform meshes.  Each mesh was 2.4~m by 1.2~m and 0.6~m tall.  The time domain for each model was chosen to be 300 seconds, which is similar to the burn time for each of the fuels.

@@

FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the source fire experiments.  The thermocouples were positioned in a vertical array, centered on the burner, beginning 0.1~m above the burner surface and spaced every 0.1~m up to 1.2~m above the burner surface.  A total heat flux gauge was also modeled to provide data at the 0.95 m from the centerline of the burner and 0.5 m above the top of the burner.  
In addition, FDS features devices that can not be physically implemented.  One such device, is used to determine the heat release rate per unit length.  This device was used to determine flame height~\cite{McDermott:2011}.

@@

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combustion zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Adjusting the FDS corresponding predictions from the RI=10 and RI=20 models down to the 0.3~m position appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explanation may be only one possibility.

These simulation results are consistent with the comparisons made to the McCaffrey's plume correlation from his study, `Purely Buoyant Diffusion Flames'~\cite{McCaffrey:1979}, in the FDS Validation Guide~\cite{FDS_Validation_Guide}. McCaffrey used a natural gas burner, 0.30~m square, which is similar in size to the burner used in this study.  McCaffrey noted that, "visually non-axisymmetric behavior could be observed only very close to the burner surface".  In the McCaffrey study, temperatures were measured with Type K thermocouples covered with a 1.0~mm stainless steel sheath. The peak heat release rate in McCaffery's plume study was 57.5~kW, this heat release rate was also the closest to the 80~kW fires in the current study.  The comparison between the McCaffrey plume correlation and the FDS simulations at RI values of 5, 10 and 20 are shown in Figure~\ref{McCaffrey_Plume_comp}. Focusing on the data between the burner surface and the 1~m above the burner surface, FDS under predicted the temperature near the surface of the burner and then over predicted as the elevation increased to approximately 0.4~m and 1.3~m above the burner for the RI=10 and 20 cases respectively. Once the temperature comparisons were above the continuous and intermittent flame regions, the agreement between the cases and the correlation began to improve.  Similar to the current study, increased RIs did not improve the simulation of the vertical centerline temperatures close to the burner.

@@

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved agreement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead may be generating the differences between the gas temperature and the measured temperature FDS values.

@@

Another examination of the plume temperature predictions was made to examine the impact of the range of radiative fractions on the FDS results.  The comparisons were made using the course grid, RI=5, and radiative fraction values of 0.20, 0.22, 0.25, and 0.30, as before. The simulations with radiative fractions between 0.2 and 0.25 were within the 2~$\sigma$ limits of the temperature measurements at each elevation.

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenario showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI= 5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

The next comparisons were made with the natural gas source fire at 4 different distances from an instrumented gypsum board wall.  Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).  The simulations will replicate the same distances.  The simulation included a vertical gypsum board wall, 1.2~m wide, that extends from the floor of the model up to the top of the domain.  The material properties of the wall are based solely on the properties of gypsum that were cited in Chapter 4.  The values that are compared in the section include: vertical centerline temperatures above the burner, wall temperatures, and heat flux values at the position of the wall surface.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.28 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500~$^\circ$C~\cite{Gross:1985}.  While FDS can accommodate changes in thermo-conductivity and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

@@

 The temperatures measured by thermocouples positioned along the vertical centerline of the natural gas burner will be compared with FDS generated temperatures from two simulations, one with RI-5 and one with RI=10.  The measurement points were at intervals of 0.2~m from 0.2~m to 2.2~m above the burner.  Figure~\ref{FDS_Plume_IWGB_comp} contains four graphs.  Each graph shows the same types of comparison, the only difference is the distance away from the gypsum wallboard.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.

@@

Figure~\ref{FDS_Wall_Temp_IWGB_comp} is analogous to Figure~\ref{FDS_Plume_IWGB_comp} except it compares the gypsum wallboard surface temperatures against the FDS simulated temperatures.  So while the FDS values for the plume over predicted the temperatures, the FDS values for the wall surface temperatures are under predicting.  This is consistent with the flame and plume moving away from the plume centerline and toward the wall.  The agreement between the FDS results and the measurements improves as the burner moves closer to the wall, however once the burner is against the wall, the degree of agreement between the FDS results and the measured temperatures decreased.  At this position the FDS results are over predicting the temperature. This differences at the wall are most distinct in the region where the flame impinged on the wall.  The agreement improves in the plume area above the flames.  In the comparisons with the source fire experiments, the FDS generated temperatures were under predicting closer to the burner surface and the over predicting in the flame region as a result of the `combustion' occurring at higher elevations above the burner.  If a similar phenomena is occurring here, the wall maybe exaggerating the effect.

@@

The instrumented gypsum wallboard had two vertical arrays of total heat flux gauges positioned at 0.2~m intervals.  Each array had six gauges installed from 0.2~m up through 1.2~m above the burner.  One array was installed in line with the center of the burner and the other array was installed on the right edge of the burner.  The comparisons between the measured heat flux and the FDS simulations are provided in Figure~\ref{FDS_HFCL_IWNG_comp} and Figure~\ref{FDS_HFEdge_IWNG_comp}.  For the centerline gauges the trends between the heat flux measurements and the FDS values are similar to the trends with the temperatures and are consistent with the flame moving or leaning toward the wall.  This resulted in higher heat flux measures relative to the FDS values at the three positions away from the wall. When the burner was positioned against the wall the FDS values and the measurements converged to a higher level of agreement.

@@

The FDS values significantly under predict the measured heat flux values along the right edge of the burner.  The best agreement occurs at the burner positions away from the wall.  At this position, the fire pattern on the walls show that the lines of demarcation are occurring very close to the edge of the burner.  Again flame movement seems to to be the main cause of the difference that is accentuated by a measurement location that is very sensitive to heat flux and position.

@@

Reviewing the videos from a natural gas burner instrumented gypsum wallboard experiment with the burner positioned at 0.310~m from the wall, we can see examples of the flame movement.  In the top row of Figure~\ref{FDS_IWNG_Flame_comp} the three images exhibit points in time where the flame was moving toward the instrumented wall and away from the vertical array of thermocouples that are positioned above the center line of the burner.  The bottom row of images are from the same fire experiment looking at it from the front.  These images also show how the pulsations and movements of the flame change positions relative to the fixed positions of the thermocouples and heat flux gauges. 
Figure~\ref{FDS_IWNG_SMV_comp} has a front view and a side view Smokeview image from the FDS RI=10 simulation with the burner positioned at 0.31~m from the wall.  This simulation is comparable to the experiment used for Figure~\ref{FDS_IWNG_Flame_comp}.  While the simulation does not demonstrate the amount of eddy generation or side to side movement from the front, the image from the side does show that FDS is simulating that the flame is moving toward the wall, although not to the extent that the actual flame did.
